The Third Bellman Equation

### 2019-01-29 paper/thirdbell.tex

\newcommand{\drho}{\!\setminus\hspace{-.75em}\rho}
\newcommand{\ztar}{z^\mathrm{tar}}
\newcommand{\Nnewt}{N_\mathrm{steps/Newt}}
\newcommand{\noise}{\epsilon}

@@

A stochastic version can be
\begin{equation}
Z\gets (1+\eta) Z-Z\tilde \Delta Z
\end{equation}
with a stochastic estimate $\tilde \Delta$ of $\Delta$ (obtained from a
transition observed in the MDP). Usually in RL we can only sample
transitions $s\to s'$ from some initial state $s$ distributed wrt some
distribution $\rho$ (which we may not control). We then parameterize
\begin{equation}
Z_{s_1s_2}\eqd \,\E_{s_2\sim \rho} z(s_1,s_2)
\end{equation}
namely, $z$ is the density of $Z$ with respect to $\rho(s_2)$.
Equivalently, in the matrix case, $Z=z\rho$ with $\rho$ the diagonal
matrix with entries $\rho$. Then the Newton update becomes
\begin{equation}
z\rho\gets 2z\rho-z\rho\Delta z\rho
\end{equation}
which is equivalent to
\begin{equation}
z\gets 2z-z\rho\Delta z
\end{equation}
and now $\rho \Delta$ is samplable:
\begin{equation}
\rho\Delta=\E_{s\sim \rho,\,s'\sim P(s'|s)}
\1_s\transp{(\1_s-\gamma\1_{s'})}
\end{equation}
so we can apply the stochastic version of the Newton update. Explicitly
this is
\begin{equation}
z(s_1,s_2)\gets (1+\eta) z(s_1,s_2)-
\E_{s\sim \rho,\,s'\sim P(s'|s)}z(s_1,s)\left(z(s,s_2)-\gamma
z(s',s_2)\right)
\end{equation}

@@

>>>>

\newcommand{\drho}{\!\setminus\hspace{-.75em}\rho}
\newcommand{\ztar}{z^\mathrm{tar}}
\newcommand{\zetatar}{\zeta^\mathrm{tar}}
\newcommand{\Nnewt}{N_\mathrm{steps/Newt}}
\newcommand{\noise}{\epsilon}

@@

A stochastic version can be
\begin{equation}
Z\gets (1+\eta) Z-Z\tilde \Delta Z
\end{equation}
with a stochastic estimate $\tilde \Delta$ of $\Delta$ (obtained from a
transition observed in the MDP). Usually in RL we can only sample
transitions $s\to s'$ from some initial state $s$ distributed wrt some
distribution $\rho$ (which we may not control). This means we can sample
from $\rho \Delta$, more precisely,
\begin{equation}
\rho\Delta=\E_{s\sim \rho,\,s'\sim P(s'|s)}
\1_s\transp{(\1_s-\gamma\1_{s'})}
\end{equation}
A solution is to parameterize
\begin{equation}
Z_{s_1s_2}\eqd \,\E_{s_2\sim \rho} z(s_1,s_2)
\end{equation}
namely, $z$ is the density of $Z$ with respect to $\rho(s_2)$.
Equivalently, in the matrix case, $Z=z\rho$ with $\rho$ the diagonal
matrix with entries $\rho$. Then the Newton update becomes
\begin{equation}
z\rho\gets 2z\rho-z\rho\Delta z\rho
\end{equation}
which is equivalent to
\begin{equation}
z\gets 2z-z\rho\Delta z
\end{equation}
and now $\rho \Delta$ is samplable,
so we can apply the stochastic version of the Newton update. Explicitly
this is
\begin{equation}
z(s_1,s_2)\gets (1+\eta) z(s_1,s_2)-
\E_{s\sim \rho,\,s'\sim P(s'|s)}z(s_1,s)\left(z(s,s_2)-\gamma
z(s',s_2)\right)
\end{equation}

@@

\paragraph{Parameterizing $Z-\Id$ in the continuous case.} I think it is quite
important to initialize at $Z=\Id$, but this is not easy in the
continuous case (initializing a neural network $z(s_1,s_2)$ to
$\delta_{s_1s_2}/\rho(s_2)$ is not convenient). Moreover, there is an
analytical singularity: for a discrete-time Markov process in continuous
space, the fundamental matrix $Z=\sum_t \gamma^t P^t$ does have two
analytically distinct components: for every $t\geq 1$ the term $P^t$ is
smooth, while for $t=0$ it is a Dirac component along the diagonal
$s_1=s_2$ (corresponding to $\Id$). So it may not be a good idea to
represent them together in the same network. Instead, we can choose a
representation of the operator $Z$ as $Z=\Id+\E_\rho \zeta$ or more
precisely
\begin{equation}
Z_{s_1s_2}=\delta_{s_1s_2}+\E_{s_2\sim \rho} \zeta(s_1,s_2)
\end{equation}
so that the operator $Z$ acting on a reward $R$ is
\begin{equation}
V(s_1)=(ZR)(s_1)=R(s_1)+\int_{s_2} \zeta(s_1,s_2)R(s_2)\d\rho(s_2)
\end{equation}

Let us work out the Newton step in this representation. Let us first go
back to the discrete case and matrices. Decomposing $Z=\Id+\zeta\rho$,
the variable $z$ is $z=Z\rho^{-1}=\rho^{-1}+\zeta$, and the objective on
$z$ is 
\begin{equation}
z\gets 2z-z\rho\Delta z=\rho^{-1}-\zeta\rho\zeta+\gamma(
P\rho^{-1}+\zeta\rho P \rho^{-1}+P\zeta+\zeta\rho P \zeta)
\end{equation}
so that the objective on $\zeta$ is
\begin{equation}
\zeta\gets -\zeta\rho\zeta+\gamma(
P\rho^{-1}+\zeta\rho P \rho^{-1}+P\zeta+\zeta\rho P \zeta)
\end{equation}
The terms $-\zeta\rho \zeta$ and $\zeta\rho
P \zeta$ are similar to what we had with $z$, but the other terms are not
obviously samplable.

The term in $P$ is a model of the transition probabilities. The term in
$P\zeta$ is the left Bellman equation, the term in $\zeta P$ is the right
Bellman equation (up to densities $\rho$ since $\zeta$ is defined as a
density). I think this is how we get the right mix of the three Bellman
equations. (Although we evaluate $P$, we do not really develop a world
model, we only need to estimate $P(s'|s)$ given $s$ and $s'$, not build
$s'$ from scratch as a sample from $P(s'|s)$.)

The non-samplability problem solves itself when considering the gradient of the
error. Let $\zetatar$ be the target above on $\zeta$. Define the loss
$\E_{s_1\sim \rho,\,s_2\sim \rho} (\zeta(s_1,s_2)-\zetatar(s_1,s_2))^2$
as above, whose gradient is
\begin{equation}
\E_{s_1\sim \rho,\,s_2\sim \rho} \,\partial \zeta(s_1,s_2)(\zeta(s_1,s_2)-\zetatar(s_1,s_2))
\end{equation}
and let us study the contributions from each term of $\zetatar$, starting
with the non-samplable term $\gamma P \rho^{-1}$. Its contribution to the
gradient is
\begin{align}
\E_{s_1\sim \rho,\,s_2\sim \rho} \,\partial \zeta(s_1,s_2)
P_{s_1s_2}/\rho(s_2)&=\E_{s_1\sim \rho}\sum_{s_2} \partial
\zeta(s_1,s_2)P_{s_1s_2}
\\&=\E_{s_1\sim \rho,\,s_2\sim P(s_2|s_1)}\partial
\zeta(s_1,s_2)
\end{align}
that is, the gradient of $\zeta$ along a transition $s_1\to s_2$.
So each time we sample a transition $s\to s'$, the term $
P\rho^{-1}$ contributes $\gamma \partial
\zeta(s,s')$ to the gradient. (This differs from the updates above, in
that above we sampled $s\to s'$ and then modified the value of $z$ at
$s_1$ and $s_2$ for independently sampled states $s_1$ and $s_2$; here
sampling $s\to s'$ contributes a change of $\zeta$ at $s$ and $s'$.)

Likewise, the gradient contribution of the $P\zeta$ term is
\begin{align}
\E_{s_1\sim \rho,\,s_2\sim \rho} \,\partial \zeta(s_1,s_2)
\,(P\zeta)(s_1,s_2)&=\E_{s_1\sim \rho,\,s_2\sim \rho}
\partial \zeta(s_1,s_2)\E_{s_3\sim P(s_3|s_1)}\zeta(s_3,s_2)
\end{align}
so this time the transition is $s_1\to s_3$. That is, when we sample a
transition $s\to s'$, we have a gradient term $\E_{s_2\sim \rho}
\partial\zeta(s,s_2)\zeta(s',s_2)$.

The last non-obvious term is $\zeta\rho P \rho^{-1}$. Its gradient
contribution is
\begin{equation}
\E_{s_1\sim \rho,\,s_2\sim \rho} \,\partial \zeta(s_1,s_2)\sum_{s_3}
\zeta(s_1,s_3)\rho(s_3)P_{s_3s_2}/\rho(s_2)
\end{equation}
because the $s_1s_2$ entry of $\zeta\rho P \rho^{-1}$ is $\sum_{s_3}
\zeta(s_1,s_3)\rho(s_3)P_{s_3s_2}/\rho(s_2)$. This rewrites as
\begin{equation}
\E_{s_1\sim \rho}\sum_{s_2} \E_{s_3\sim\rho} P_{s_3s_2}
\zeta(s_1,s_3)\partial\zeta(s_1,s_2)
\end{equation}
which in turn rewrites as
\begin{equation}
\E_{s_1\sim \rho}\E_{s_3\sim\rho} \E_{s_2\sim P(s_2|s_3)}
\zeta(s_1,s_3)\partial\zeta(s_1,s_2)
\end{equation}
that is, when we sample a transition $s\to s'$, we have a contribution
$\E_{s_1\sim \rho}\zeta(s_1,s)\partial \zeta(s_1,s')$.

Thus, when sampling $s\to s'$, we have several gradient contributions:
one is similar to what we have before and involves
$\partial\zeta(s_1,s_2)$ for independent samples $s_1,s_2$; one involves
$\partial\zeta(s,s')$, one involves $\partial\zeta(s,s_2)$, and the last
involves $\partial\zeta(s_1,s')$. All are samplable.

TODO rewrite gradient with current $\zeta$ and previous $\zeta_0$, with
all correct signs and $\gamma$ factors...

% Starting with a
% previous value $Z_0=\Id+\E_{\rho}\zeta_0$, or
% $z_0(s_1,s_2)=\delta_{s_1s_2}/\rho(s_2)+\zeta_0(s_1,s_2)$, the target is
% $2Z_0-Z_0\Delta Z_0$. The loss is $\E_{s_1\sim \rho,\,s_2\sim \rho}
% \,(z(s_1,s_2)-\ztar(s_1,s_2))^2$, where $z$ and $\ztar$ are the density
% of $Z$ and of the target with respect to $\rho(s_2)$. On the variable
% $z$, the target $\ztar$ is
% $\ztar=2z_0-z_0\rho\Delta z_0$ which rewrites as
% \begin{equation}
% \ztar(s_1,s_2)=2\frac{\delta_{s_1s_2}}{\rho(s_2)}+2\zeta(s_1,s_2)-(\frac{\delta_{s_1s_2}}{\rho(s_2)}+\ztar(s_1,s_2))...
% \end{equation}
% % The expected
% % gradient of the loss is $2\E_{s_1\sim \rho,\,s_2\sim \rho}\partial
% % z(s_1,s_2)(z(s_1,s_2)-\ztar(s_1,s_2))$. We find
% % \begin{align}
% % z(s_1,s_2)-\ztar(s_1,s_2)=\frac{\delta_{s_1s_2}}{\rho(s_2)}-2\frac{\delta_{s_1s_2}}{\rho(s_2)}-2\zeta_0(s_1,s_2)+
% %\end{align}
